---
title: "fishSim-vignette"
author: "Shane M Baylis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fishSim-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# fishSim vignette
## built `r Sys.Date()`.

fishSim is a package for demographic simulation and exploration of kin relationships. It includes functions to build populations and move, mate, kill, age, and switch the sex of individuals within those populations or defined subpopulations. Importantly, fishSim is individual-based and retains parentage information for all individuals, allowing full reconstruction of kin relationships to arbitrary depth.

In addition to population simulation features, fishSim includes a set of convenience functions allowing the user to check the scenario population growth rate, find simulation scenarios with null population growth (`PoNG()`), selectively archive subsets of the population for increased computational speed. For a given population, fishSim also includes functions to look up pairs of animals, compare their sets of ancestors, and classify them into kin categories based on their nearest shared ancestor(s), or show patterns of shared ancestors for members of a pair.

fishSim is developed in response to the demands of close-kin mark-recapture projects. 'typical' use-cases therefore involve an understanding of both demography and kin relationships. This vignette will cover two use scenarios - one relatively simple one, with:

- age-specific (but not sex-specific) fecundity,
- constant adult survival,
- a single population with no subdivisions (i.e., one 'stock'),
- no sex-switching, and
- first-year survival set by `PoNG()`, giving a flat population size through time,

...and a one designed to show off the bells and whistles, with:

- age-specific, sex-specific fecundity with within-season mating exhaustion for females,
- age-specific, sex-specific survival,
- three subpopulations ('stocks') with markovian movement between subpopulations,
- male-to-female sex-switching,
- a growing population.

In the 'bells-and-whistles' scenario, we will also cover archiving tools. For both scenarios, we will take a brief look at relationship patterns between pairs of animals.


## The relatively-simple scenario

### Setup

In order to simulate a population, we need an object to hold that population's data, and the population will need some founding members. Our data object will have one line per individual, so let's call it `indiv`.

```{r}
library(fishSim)
indiv <- makeFounders(stocks = c(1))
head(indiv)
```

Now, what's here? `indiv` has one line per individual, and each individual has an ID code (`indiv[,1]`), a sex (`indiv[,2]`), a father's ID code (`indiv[,3]` - but which is 'founder' here, because founding animals can't really be said to have parents), a mother's ID code (`indiv[,4]` - ditto), a birth year `indiv[,5]`, a death year (`indiv[,6]` - NA for any animals that are still alive), a stock membership (`indiv[,7]`), and an age in years (`indiv[,8]`).

The age structure, sex ratio, and stock memberships of founders can all be specified in the `makeFounders()` call. We've only changed one of the default values - instead of the default 3 stocks, we're going to have a single panmictic founder population. The other defaults give us 1000 founders, with an even sex ratio, with an age-structure that implies 70% annual survival, but with a hard age-limit of 20 years.

### Mating

Having set up our population, we can now get its members to breed, get older, switch sex and die, using the `mate()` (or `altMate()`), `birthdays()`, `sexSwitch()`, and `mort()` functions. If we had more than one stock, we could also get individuals to move between stocks with markovian movement probabilities, using `move()`. Let's look at those.

```{r}
nrow(indiv)
indiv <- mate(indiv, year = 1)
nrow(indiv) ## 200 newborns
tail(indiv) ## newborns are added to the end
```

Here are some newborns, generated using the default `mate()` settings. Unlike founders, these newborns all have a father's ID and a mother's ID, are all aged 0, and all have a birth year of equal to the `year` argument. There are 200 newborns, because `mate()` generates new members as a proportion of the extant population size (i.e., we specify fecundity for the population, not the individuals), and that proportion (argument `fecundity`) is 0.2 by default. Each individual was born in a clutch, and each clutch contained a Poisson-distributed number of clutch-members, the default mean ('batchSize') of 0.5. `mate()` simply kept producing newborns by mating together random male/female pairs until `nrow(indiv) * 0.2` new offspring were produced.

Alternatively, we can mate our population using `altMate()`, which specifies individual maturities by age, and a probability distribution of number of offspring per mature female. Let's generate some new founders and try that.

```{r}
indiv <- makeFounders(stocks = c(1))
nrow(indiv) ## 1000 founders
indiv <- altMate(indiv, firstBreed = 2, batchSize = 0.9, year = 1)
tail(indiv)
```

In this call to `altMate()`, females became sexually mature at 2 years of age (`firstBreed = 2`), and each sexually-mature female produced a Poisson-distributed number of offspring with mean set by `batchSize`. It is possible to specify age-specific or sex-specific fecundities, set paternity within-clutch to be single or multiple, to exhaust fathers within each breeding season (females, in this system, always breed to exhaustion each season), and to set a sex ratio for offspring by additional arguments to `altMate()`, and these options can also be set in `mate()`.

### Mortality

Now, let's kill some of the population. Mortality probabilities can be flat, age-specific or stock-specific (and with a little extra effort, sex- or age:stock:sex-specific too), or we can randomly kill animals such that the population is reduced to a certain size, and we can set an age past which no animal will survive. For now, let's set a flat 20% mortality rate (i.e., the probability of death is equal for all animals).

```{r}
nrow(indiv)
indiv <- mort(indiv, year = 1, type = "flat", mortRate = 0.2)
nrow(indiv)
head(indiv, n = 15)
```

Mortality does not remove dead animals from 'indiv' - we will need to refer to them later - it just updates their death year to the value given in `year`. In general, though, animals with a non-NA value for death year will not move, mate, switch sex, or have birthdays.

### Birthdays

On the topic of birthdays: that's the final thing that must be done before we can put this all together and run a full (albeit basic) demographic simulation: we need to be able to increase the age of our animals, and that is what the `birthdays()` function is for. It's very simple: it takes all living members of the population, and increments their age by 1.

```{r}
tail(indiv)
indiv <- birthdays(indiv)
tail(indiv)
```

### What's my population doing? Can I make it stay the same size?

A couple more functions deserve a mention here: `check_growthrate()` and `PoNG()`. `check_growthrate()` tells you how quickly your population is growing. `PoNG()` tells you what you need first-year survival to be, in order for your population to maintain a constant size (within limits - the estimation is Leslie Matrix-based, and there are ways to fool Leslie Matrices that are out of scope for this vignette but covered briefly in the `PoNG()` and `check_growthrate()` documentation). Let's try both of those, using the mating and mortality settings we have already used.

```{r}
check_growthrate(mateType = "flat", mortType = "flat",
                 batchSize = 0.9, firstBreed = 2, mortRate = 0.2)

## with the current settings, we expect our population to grow by about 1.8% annually.

PoNG(mateType = "flat", mortType = "flat", batchSize = 0.9, firstBreed = 2,
     mortRate = 0.2)

## if first-year mortality was about 0.305, rather than 0.2, our population
## would have null growth. You can also read off a range of possible growth rates
## from the plot.
```

### Turning processes into a simple simulation

To turn those processes into a full demographic simulation, all that is needed is to repeat the processes in a loop. Let's do that, but set age-specific mortality so that the long-run average population growth rate is zero using the first-year mortality rate we got from `PoNG()`, so that our population is unlikely to explode or become extinct.

```{r}
indiv <- makeFounders(stocks = c(1))
ageMort <- c(0.305, rep(0.2, 100)) ## age-specific mortality is 0.305 for first-years,
                                   ## 0.2 for all older age-classes.

for (y in 1:60) {
    indiv <- altMate(indiv, firstBreed = 2, batchSize = 0.9, year = y) ## y for year
    indiv <- mort(indiv, year = y, type = "age", ageMort = ageMort)    ## age-specific mort
    indiv <- birthdays(indiv)
}
tail(indiv) ## a population with 60 years of births, deaths, and birthdays
nrow(indiv[is.na(indiv[,6]),]) ## the currently-alive population size. Note that population
                               ## growth only *averages* zero, and variability occurs!
```

### Looking up relationships between pairs of animals

One of the key advantages of fishSim is its ability to report on different relationships between pairs of animals in a sample. Are these two each other's siblings? Half-siblings? Is one the parent of the other? The grandparent? Some obscure half-cousin, once removed?


```{r}
## take a sample of 100 individuals born in the last 10 years of the simulation
sampled <- sample(indiv[indiv[,5] > 50,][,1] , size = 100) 

## look up each animal's ancestors and look for shared ancestors between each
## pair of animals:
pairs <- findRelativesPar(indiv = indiv, sampled = sampled)

HSPs <- pairs[pairs$TwoTwo == 1,] ## pairs that are half-siblings of each other

## look at the number of shared ancestors at each ancestral
## generation, for one of the half-sibling pairs.
lookAtPair(HSPs[1,]) 

relatives <- namedRelatives(pairs) ## shows the number of pairs of each relationship type
relatives
```

## The 'bells and whistles' scenario

In this second scenario, we will simulate a metapopulation with:

- age-specific, sex-specific fecundity with within-season mating exhaustion for females,
- stock-specific, age-specific survival,
- three subpopulations ('stocks') with markovian movement between subpopulations,
- male-to-female sex-switching,
- a growing population.
- archiving for speed (?)

We set up the population basically as before, with a couple of slight tweaks. First, we should make sure the founding population has multiple stocks (we'll give it three). Second, we will set up an archive matrix. The archive matrix is intended to hold the records of dead animals in large, long-running simulations: dead animals do not take part in any further mating, movement, sex-switching, aging, or mortality events, but if they stay in `indiv`, `indiv` can become huge and unweildy, slowing down all of those processes. There is of course a trade-off, in that writing dead animals to the archive takes system time, so the optimum may be to only archive dead animals once every few 'years'.

```{r}
## set up founders with three stocks: two that each contain 30% of the population,
## and one that contains the remaining 40%.

indiv <- makeFounders(pop = 1000, stocks = c(0.3, 0.3, 0.4))

## set up archive - just a matrix with zero rows and eight columns

archive <- make_archive()

```

Because we will have inter-stock movement in this sim, we will also need to set up a matrix giving the probability that an individual will move into another stock, given its current stock membership. Because survival will be age- and stock-dependent, we will need a matrix of survival rates with as many columns as stocks, and as many rows as (at least) the age of the oldest plausible population-member. Because we will have age-specific, sex-specific fecundity, we will need a male age-specific maturity curve and a female age-specific maturity curve. Let's set those up now.

```{r}
## Markovian movement matrix
stocks <- c(0.3, 0.3, 0.4)
admix.m <- matrix(NA, nrow = length(stocks), ncol = length(stocks))
for(i in 1:nrow(admix.m)) {
    admix.m[i,] <- stocks*stocks[i]
}
## admix.m shows movement proportional to starting population sizes.
admix.m
## let's tweak those numbers so that animals tend to stay where
## they are, and not move around so much.
admix.m <- matrix(c(0.23, 0.03, 0.04, 0.03, 0.23, 0.04, 0.04, 0.04, 0.32),
                  nrow = length(stocks), ncol = length(stocks), byrow = FALSE)
admix.m

## Age- and stock-dependent survival
ageStockMort <- matrix(c(0.47, 0.37, 0.27, rep(0.23, 97),
                         0.45, 0.35, 0.25, 0.20, rep(0.22, 96),
			 0.45, 0.3, 0.3, 0.19, rep(0.2, 96)),
                       ncol = length(stocks), nrow = 100)
head(ageStockMort)

## Sex-specific maturity curves

maleCurve <- c(0,0,0,0.1,0.5,0.8,0.85,0.9,0.95,rep(1, 91))
femaleCurve <- c(0,0,0.5,0.9,0.95,rep(1,95))
## maleCurve and femaleCurve should both be long enough that no individuals
## will outlive the curve.
head(maleCurve)
head(femaleCurve)

```

Now, what's going to happen with these subpopulations, if we run with those parameters?

```{r}
check_growthrate(mateType = "ageSex", mortType = "ageStock", batchSize = 1.6,
                 femaleCurve = femaleCurve,
                 ageStockMort = ageStockMort)
## Not bad. Two of the three populations are increasing. The third will probably be kept viable
## by immigration from the other two. Note that I fiddled 'batchSize' (which is the mean number
## of offspring per mature female per breeding attempt) a bit.
```

Now we just run the population simulator, as we did for the simple population, but with a couple of extra things to specify. Let's give it 60 years.


```{r}
for (k in 1:60) {
    ## very rarely, switch some males to females
    indiv <- sexSwitch(indiv = indiv, direction = "MF", prob = 1e-04)
    ## move animals according to the markovian matrix we set up before
    indiv <- move(indiv = indiv, moveMat = admix.m)
    ## mate animals using the age-specific, sex-specific curves we set up before
    indiv <- altMate(indiv = indiv, batchSize = 1.6, type = "ageSex", maleCurve = maleCurve,
                     femaleCurve = femaleCurve, year = k)
    ## kill animals on the basis of their age and stock, as set up before   
    indiv <- birthdays(indiv = indiv) 
    indiv <- mort(indiv = indiv, year = k, type = "ageStock", ageStockMort = ageStockMort)
    if(k %in% c(10,20,30,40,50,60)) {
        archive <- archive_dead(indiv = indiv, archive = archive)
	indiv <- remove_dead(indiv = indiv)
	print(paste("Iteration ", k, " complete", sep = ""))
	## do some clean-up, but only occasionally. And report on progress.
    }
}
archive <- rbind(archive, indiv)
indiv <- archive  ## merge 'indiv' and 'archive', since they were only separated for speed.

```

Now that the simulation is finished, examining kin relationships is *exactly* the same as the simple scenario:

```{r}
## take a sample of 100 individuals born in the last 10 years of the simulation
sampled <- sample(indiv[indiv[,5] > 50,][,1] , size = 100) 

## look up each animal's ancestors and look for shared ancestors between each
## pair of animals:
pairs <- findRelativesPar(indiv = indiv, sampled = sampled)

HSPs <- pairs[pairs$TwoTwo == 1,] ## pairs that are half-siblings of each other

## look at the number of shared ancestors at each ancestral
## generation, for one of the half-sibling pairs.
lookAtPair(HSPs[1,]) 

relatives <- namedRelatives(pairs) ## shows the number of pairs of each relationship type
relatives
```
